---
title: "Allies and Enemies"
output: html_document
---

**This homework is due Sunday April 10, 2016 at 11:59PM EST. When complete, submit your code in an R Markdown file and the knitted HTML via GitHub.**

# Introduction 

The United Nations (UN) is an intergovernmental organization 
founded in 1946 to promote international cooperation. It now 
represents 193 member states. The General Assembly is the largest 
body, with a seat for every member of the UN. It discusses 
topics of international importance such as maintaining peace 
and security, providing humanitarian aid, and protecting human rights. 

We will be analyzing a dataset containing the full history of 
General Assembly votes by each country to determine what 
countries vote similarly and which do not. We will also 
explore how this changes through time.


# Problem 1

We'll start by loading the United Nations voting data into R 
and performing some data wrangling. We use data from this paper:

> Voeten, Erik; Strezhnev, Anton; Bailey, Michael, 2009, "United Nations General Assembly Voting Data", http://hdl.handle.net/1902.1/12379, Harvard Dataverse, V11

In this problem, we will combine information from three sources 
to create the datasets that we will use to study voting behavior.

### Problem 1A

We have learned how to import text files into R. Here we are 
going to load a data object that is saved to a file. To get 
an idea of how this works try the following:

```{r}
temp_filename <- tempfile() ## creaate tempory file name
temp_object <- 1:5 ## create an R object
save(temp_object, file=temp_filename) ## save the r object to file
rm(temp_object) ## remove object
load(temp_filename) ## load object from file
temp_object ## note that it's back
```

We usually use the suffix `.RData` or `.rda` for these objects. 

The data for this project is stored as an `.RData` file. Go to 
[this web page](https://dataverse.harvard.edu/dataset.xhtml?persistentId=hdl:1902.1/12379). 
To get the `.RData` file, click on the Download button for 
`rawvotingdata13.tab` and choose the `RData format`.

To load the data set into R, use the `load()` function.
Define the name of the object as `x` (but do NOT print it 
out as it has over 1 million rows).

```{r}
## put your code here

```


### Problem 1B

The first problem to overcome is that if you try to print 
this object, it will crash your R session -- it's just that 
big! (`r nrow(x)` rows). So first wrap it in `tbl_df(x)`,
and call it `votes`. After doing this you can erase `x` with `rm(x)`.

```{r}
## put your code here

```


### Problem 1C

We note that the data is already arranged according to the 
rules of tidy data. There is one row for each observation 
and one column for each variable. 

Download the `Codebook.pdf` file from [this page](https://dataverse.harvard.edu/dataset.xhtml?persistentId=hdl:1902.1/12379). 
How would you interpret the vote column? How many of each 
kind of vote are in this dataset? 

```{r}
## put your code here

```


Of the five types of votes, which three would provide information 
about the country's position on an issue? Which two would not?

**Your answer here**: 


Filter out the types of votes that do not provide information 
about our countries position on an issue from our dataset.

```{r}
## put your code here

```


### Problem 1D 

According to the codebook, which column represents countries? 
What type of unique code is used to represent each country?

**Your answer here**:  

Create new `country` column that contains country names 
based on this column. Hint: check out the [countrycode](https://cran.r-project.org/web/packages/countrycode/countrycode.pdf) 
package. 


```{r}
## put your code here

```


### Problem 1E

Before continuing let's wrangle the country names a bit. We are 
renaming countries with long names and renaming Congo to 
distinguish it from Democratic Republic of Congo. We make use 
of the powerful remapping function `revalue()` from `plyr` 
package. You should **not** load `plyr` though as it will 
create confusion with `dplyr` functions.

```{r}
library(tidyr)
mapping <- c("United States"="USA",
          "United Kingdom"="UK",
          "Korea, Republic of"="South Korea",
          "Lao People's Democratic Republic"="Laos",
          "Yemen People's Republic"="South Yemen",
          "Saint Vincent and the Grenadines"="Saint Vincent",
          "Congo"="Congo Republic")
votes <- votes %>% mutate(country = plyr::revalue(country, mapping)) %>%
  separate(country, into = c("country", "extra"), sep=",", fill="right")
```

Right now we have information about how every country voted 
on every resolution. But we do not have any information about
the resolutions themselves (e.g. not what their title or topic
was, or what date they were voted on). 
Next, we will bring this data in as well.

This data is provided as `descriptions.csv`. 
Read it in using the `readr` package and wrangle it as shown below:

```{r}
library(readr)

url <- "https://raw.githubusercontent.com/datasciencelabs/data/master/un-resolutions-descriptions.csv"
descriptions <- read_csv(url, col_types = list(date = col_date("%m/%d/%y")))

## from warning and looking at csv we see
## line 1483 has an extra "
## it's supposed to be a 0
descriptions[1483,"ec"] <-0

library(lubridate)
y <- year(descriptions$date)
year(descriptions$date) <- ifelse(y > 2030, y - 100, y)
```

Count the number of votes that were taken in each year. 
Create a line graph of the number of votes per year.

```{r}
## put your code here

```


What year would we want to filter out from the dataset 
because there was only one vote? 

**Your answer here**: 

Filter it out now.

```{r}
## put your code here

```


### Problem 1F

Read the `Codebook.pdf` about this dataset. Who classified certain votes 
as "important"? 

**Your answer here**:  

What percent of votes in history were categorized as "important"?

```{r}
## put your code here

```


The most interesting analyses can be done by combining the description 
and country-voting data.

Join the `description` and country-voting data (`votes`) to create a new 
data set. Remove the `yes`, `no`, and `abstain` columns from the 
`description` dataset. These are per-vote summaries that we do not 
need any more (and could be misleading). The final dataset should be called 
`votes`, which you will continue to use throughout the homework. 

```{r}
## put your code here

```



# Problem 2

### Problem 2A

Canada and the US have been allies since the UN was created. 
We can create a matrix of all votes for these two countries using 
the `spread()` function in `tidyr` package like this:

```{r}
library(tidyr)
y <- votes %>% 
  filter(country %in% c("USA", "Canada")) %>%
  mutate(year = year(date)) %>%
  select(rcid, year, importantvote, country, vote) %>%
  spread(country, vote)
```

We can see how often they have voted together in important votes 
and not-important votes:

```{r}
y %>% 
    group_by(importantvote) %>% 
    summarize(mean(USA==Canada, na.rm=TRUE))
```

Compute the percentage in which the US and Canada
voted the same. Calculate this percentage for each year and call it 
`agreement`. Fit a linear model using `lm()` to predict `agreement`
with `year`. 

```{r}
## put your code here

```

What is the trend predicted by the linear model? 
Is it statistically significant?

**Your answer here**: 



### Problem 2B

In the previous problem we found a negative trend in the agreement 
between the USA and Canada throughout the years. Interpreting this 
linear model would imply that disagreement between these two counties 
was worse during the Clinton administration (1992-2000) than the 
Reagan administration (1980-1998). 

Now, instead of blindly interpreting the regression results, 
plot the data and use a smoother to estimate a trend. Based on this 
analysis, how do thes Regan and Clinton administrations compare? 

**Hint**: Make sure to pick a window size or span that creates 
a trend that goes through data.

```{r}
## put your code here

```


### Problem 2C

Make the plot above for the agreement through time between the US
and the following countries: Israel, UK, Mexico, Cuba, and China. 
Make two plots: one for important votes to non-important votes. 

```{r}
## put your code here

```

Describe the observed patterns.

**Your answer here**: 



# Problem 3

In this problem, we will focus only on important votes. 
To get a better idea of who votes together we can compute a 
distance between each country. We will focus on countries that 
voted more than 95% of time in the 368 votes

```{r}
countries <- votes %>% 
                filter(importantvote==1) %>% 
                group_by(country) %>% 
                summarize(p=n()/368) %>% 
                filter(p>=0.95) %>% 
                .$country
```

We can create a matrix with all the votes using the `spread()` function:

```{r}
tmp <- votes %>% 
    filter(country %in% countries & year(date) >= 1980 & importantvote == 1) %>%
    select(rcid, country, vote) %>% 
    spread(country, vote) 

X <- as.matrix(tmp[,-1])
rownames(X) <- tmp$rcid
```

### Problem 3A

Create a distance matrix between each country. Call this matrix `d`. 

**Hint**: Use the `dist()` function, but note that `X` has 
countries in the columns and `dist()` computes distances between rows. 
Look at the `dist` help file for more infomration. 
You can use the default `method = "Euclidean"` in the `dist()` function. 
You can switch rows to columns using the `t()` (transpose) function. 
Finally, once you create the distance matrix `d` you can 
visualize it using `heatmap()` or `hclust()`.

```{r}
## put your code here

```


What country is closest to US? Which is furthest?

**Your answer here**: 


### Problem 3B

Given how close some countries are and how far others are to US in voting,
we should be able to predict how the US will vote based on others. 
Let's try to implement a machine learning algorithm to do this. 

Use the `votes` data set to create a new dataset with seven columns. 
One column will represent the USA vote as the outcome (call it `y`) and
the last six columns will be the vote from the six countries examined 
above in Problem 2 (include Canada), which will be used a predictors 
in our machine learning algorithm. Only consider the important votes. 
In the column for the USA vote column (`y`), remove the `Abstain` votes and
only consider the `Yes` and `No` votes from the USA. Tranform the USA vote
column (`y`) to contain only 0s and 1s where 0 = `No` vote and 1 = `Yes` vote.

```{r}
## put your code here

```

Use the `caret` R package to split the data into a training set with 
80% of data and a test set with the remaing 20%. 
Then use `glm()` to build a model. What is the accuracy?

```{r}
## put your code here

```



### Problem 3C 

We see that obtain a very high accuracy, but note that this is a 
random variable due to the random split of our data. 
Try 10 new random splits and report on how much our accuracy changes.

```{r}
## put your code here

```


### Problem 3D 

Compare your `glm()` model to a `knn()`. Use the `train()` function 
to run 10 cross validations with leaving out 80% of the data. 
Plot your results. 

```{r}
## put your code here

```

How many nearest neighbors should we use?

**Your answer here**: 
