---
output: html_document
---

```{r, message=FALSE, warning=FALSE, echo=FALSE}
library(dplyr)
library(ggplot2)
theme_set(theme_bw(base_size = 16))
library(gganimate)
library(readr)
library(knitr)
library(broom)
library(gridExtra)
```

## Review

Let's continue with the digits data. We read-in the data:

```{r}
url <- "https://raw.githubusercontent.com/datasciencelabs/data/master/hand-written-digits-train.csv"
if(!exists("digits")) digits <- read_csv(url)
```
To simplify the problem we will try to distinguish 2s from 7s. So we subset to only those digits

```{r}
dat <- digits %>% filter(label%in%c(2,7))
```


For illustrative purposes we created two features: `X_1` is the percent of black pixels that are in the top left quadrant and `X_2` is the percent of black pixels that are in the bottom left quadrant:

```{r, echo=FALSE, fig.align="center", fig.width=14}
tmp <- lapply( c(37,9,5,28), function(i){
    expand.grid(Row=1:28, Column=1:28) %>%  
      mutate(id=i, label=dat$label[i], id=paste("obs",i),
             value = unlist(select(dat,pixel0:pixel783)[i,]))
})
tmp <- Reduce(rbind, tmp)

tmp %>% ggplot(aes(Row, Column, fill=value)) + 
  geom_raster() + 
  scale_y_reverse() +
  scale_fill_gradient(low="white", high="black") +
  geom_vline(xintercept = 14.5) +
  geom_hline(yintercept = 14.5)  +
  facet_grid(.~id) 
```

We can create these new predictors like this:

```{r, echo=FALSE, cache=TRUE}
##Get the truth
dat <- mutate(dat, label =  as.character(label)) %>% 
  mutate(y = ifelse(label=="2",0,1 ))
row_column <- expand.grid(row=1:28, col=1:28)
ind1 <- which(row_column$col <= 14 & row_column$row <=14)
ind2 <- which(row_column$col > 14 & row_column$row > 14)
ind <- c(ind1,ind2)
X <- as.matrix(dat[,-1])
X <- X>200
X1 <- rowSums(X[,ind1])/rowSums(X)
X2 <- rowSums(X[,ind2])/rowSums(X)
dat <- mutate(dat, X_1 = X1, X_2 = X2)
```

We can see some example of what these predictors are:

```{r, echo=FALSE, fig.align="center", fig.width=14}
tmp <- lapply( c(37,9,5,28), function(i){
    expand.grid(Row=1:28, Column=1:28) %>%  
      mutate(id=i, label=paste0("obs ",i,": X_1 = ",round(dat$X_1[i],2),", X_2 = ",round(dat$X_2[i],2)),
             value = unlist(select(dat,pixel0:pixel783)[i,]))
})
tmp <- Reduce(rbind, tmp)

tmp %>% ggplot(aes(Row, Column, fill=value)) + 
  geom_point(pch=21,cex=3) + 
  scale_y_reverse() +
  scale_fill_gradient(low="white", high="black") +
  geom_vline(xintercept = 14.5) +
  geom_hline(yintercept = 14.5)  +
  facet_grid(.~label) 
```  

We act as if we know the truth:

```{r, echo=FALSE}
y <- as.factor(dat$label)
x <- cbind(X1, X2)
library(caret)
fit <- knn3(x, y, 51)
GS <- 150
X1s <- seq(min(X1),max(X1),len=GS)
X2s <- seq(min(X2),max(X2),len=GS)
true_f <- expand.grid(X_1=X1s, X_2=X2s)
yhat <- predict(fit, newdata = true_f, type="prob")[,2]
true_f <- mutate(true_f, yhat=yhat)
f <- loess(yhat~X_1*X_2, data=true_f, 
           degree=1, span=1/5)$fitted
true_f <- true_f %>% mutate(f=f) 
rm(X,X1,X2,fit,GS,X1s,X2s,yhat,f)

true_f_plot <- true_f %>%
  ggplot(aes(X_1, X_2, fill=f))  +
  scale_fill_gradientn(colors=c("#F8766D","white","#00BFC4")) + geom_raster()  + #guides(fill=FALSE) +  
  stat_contour(aes(x=X_1,y=X_2,z=f), data=true_f, breaks=c(0.5),color="black",lwd=1.5)
true_f_plot
```

## Quadratic and Linear Discriminant Analysis

For illustration purposes let's take a subset:
```{r}
##create the training set with 1000 variabes
set.seed(1)
dat <- sample_n(dat, 1000) %>% select(y, X_1, X_2)
dat
```

Now create train and test sets:
```{r}
library(caret)
inTrain <- createDataPartition(y = dat$y, p=0.5)
train_set <- slice(dat, inTrain$Resample1)
test_set <- slice(dat, -inTrain$Resample1)
```


Quadratic Discriminant Analysis (QDA) relates to the _Naive Bayes_ approach we described earlier. We try to estimate 
$\mbox{Pr}(Y=1|X=x)$ using Bayes theorem. 

$$
f(x) = \mbox{Pr}(Y=1|\mathbf{X}=\mathbf{x}) = \frac{\pi p_{\mathbf{X}|Y=1}(\mathbf{x})}
{(1-\pi) p_{\mathbf{X}|Y=0}(x) + \pi p_{\mathbf{X}|Y=1}(x)}
$$

With QDA we assume that the distributions $p_{\mathbf{X}|Y=1}(x)}$ and $p_{\mathbf{X}|Y=0}(\mathbf{x})$ are multivariate normal. In our case we have two predictors so we assume each one is bivariate normal. This implies we need to estimate two averages, two standard deviations, and a correlation for each case $Y=1$ and $Y=0$. 

This implies that we can approximate the distributions $p_{X_1,X_2|Y=1}$ and $p_{X_1, X_2|Y=0}$. We can easily estimate parameters from the data:

```{r}
options(digits = 2)
params <- train_set %>% group_by(y) %>% 
  summarize(avg_1 = mean(X_1), avg_2 = mean(X_2), sd_1= sd(X_1), sd_2 = sd(X_2), r = cor(X_1,X_2))
params
```

So here are the data and contour plots showing the two normal densities:

```{r}
train_set %>% mutate(y = factor(y)) %>% 
  ggplot(aes(X_1, X_2, fill = y, color=y)) + 
  geom_point(pch=21,cex=5, color="black") + 
  stat_ellipse(lwd=2, type="norm")
```

This defines the following estimate of $f(x_1, x_2)$ 

```{r}
library(mvtnorm)

get_p <- function(params, data){
  dmvnorm( cbind(data$X_1, data$X_2), 
               mean = c(params$avg_1, params$avg_2), 
               sigma = matrix( c(params$sd_1^2,
                                 params$sd_1*params$sd_2*params$r,
                                 params$sd_1*params$sd_2*params$r,
                                 params$sd_2^2),2,2))
}
pi <- 0.5
p0 <- get_p(params[1,], true_f)
p1 <- get_p(params[2,], true_f)

f_hat_qda <- pi*p1/(pi*p1 + (1-pi)*p0)

p <-true_f %>% mutate(f=f_hat_qda) %>%
 ggplot(aes(X_1, X_2, fill=f))  +
  scale_fill_gradientn(colors=c("#F8766D","white","#00BFC4")) + 
  geom_raster()  + #guides(fill=FALSE) +  
  stat_contour(aes(x=X_1,y=X_2,z=f), 
               data=mutate(true_f, f=f_hat_qda),
               breaks=c(0.5),color="black",lwd=1.5)

grid.arrange(true_f_plot, p, nrow=1)
```


Here we have 2 predictors and had to compute 4 means, 4 SDs and 2 correlations. How many parameters would we have if instead of 2 predictors we had 10? 

The main problems comes from estimating correlations for 10 of predictors. With 10, we have 45 correlations for each class. In general the formula is $p(p-1)/2$ which gets big fast. 

A relatively solution to this is to assume that the correlation structure is the same for all classes. Which reduces the number of parameters we need to estimate. When we do this, we can show mathematical that the solution is "linear", in the linear algebra sense and we call it Linear Discriminant Analysis (LDA). 

```{r}
params <- train_set %>% group_by(y) %>% 
  summarize(avg_1 = mean(X_1), avg_2 = mean(X_2), sd_1= sd(X_1), sd_2 = sd(X_2), r = cor(X_1,X_2))
params <-params %>% mutate(sd_1 = mean(sd_1), sd_2=mean(sd_1), r=mean(r))
params 
```

This defines the following estimate of $f(x_1, x_2)$  and the boundary becomes linear:

```{r}
library(mvtnorm)
p0 <- get_p(params[1,], data=true_f)
p1 <- get_p(params[2,], data=true_f)
p <- 0.5

f_hat_lda <- pi*p1/(pi*p1 + (1-pi)*p0)

p <- true_f %>% mutate(f=f_hat_lda) %>%
 ggplot(aes(X_1, X_2, fill=f))  +
  scale_fill_gradientn(colors=c("#F8766D","white","#00BFC4")) +
  geom_raster()  + #guides(fill=FALSE) +  
  stat_contour(aes(x=X_1,y=X_2,z=f), 
               data=mutate(true_f, f=f_hat_lda), 
               breaks=c(0.5),color="black",lwd=1.5)

grid.arrange(true_f_plot, p, nrow=1)
```


### Connection distance

The normal density is

$$
p(x) = \frac{1}{\sqrt{2\pi} \sigma} \exp\left\{ - \frac{(x-\mu)^2}{\sigma^2}\right\}$$

Note if we remove the constant $1/(\sqrt{2\pi} \sigma)$ and then take the log we get:

$$
- \frac{(x-\mu)^2}{\sigma^2}
$$

which is the negative of a distance squared scaled by the standard deviation. For higher dimensions the same is true expect the scaling is more complex and involved correlations.

## ROC

With the example we have been examining we can make two types of errors: calling a 2 a 7 or calling a 7 a 2. More generally, we binary data we call these false positives (calling a 0 a 1) and false negatives (calling a 1 a 0). Here we have arbitrarily made 7s 1s and 2s 0s.  

This concept is important in many areas and in particular in health where one type of mistake can be much more costly than another. Note that we have been predicting 1s based on the rule $\hat{f}(x_1, x_2) > 0.5$ but we pick another cutoff, depending on how the cost of errors. For example, if we are predicting if a plane will malfunction, then we want a very low false negative rate and are willing to sacrifice our true positive rate. 

We can see that the estimated probabilities are in a continuum:


```{r}
params <- train_set %>% group_by(y) %>% 
  summarize(avg_1 = mean(X_1), avg_2 = mean(X_2), 
            sd_1= sd(X_1), sd_2 = sd(X_2), 
            r = cor(X_1,X_2))
p0 <- get_p(params[1,], dat= test_set)
p1 <- get_p(params[2,], dat= test_set)
pi <- 0.5
pred_qda <- pi*p1/(pi*p1 + (1-pi)*p0)
test_set %>% mutate(pred=pred_qda, label=as.factor(y)) %>%
  ggplot(aes(label,pred)) + geom_boxplot()
```

The Receiver Operator Characteristic Curve plots true positive rate versus false positive rates for several choices of cutoff. We can create this curve with the following code:

```{r}
library(pROC)

roc_qda <- roc(test_set$y, pred_qda)
plot(roc_qda)
```

Here are the results for LDA

```{r}
params <-params %>% mutate(sd_1 = mean(sd_1), 
                           sd_2 = mean(sd_1), 
                           r=mean(r))
p0 <- get_p(params[1,], dat = test_set)
p1 <- get_p(params[2,], dat = test_set)
pi <- 0.5
pred_lda <- pi*p1/(pi*p1 + (1-pi)*p0)

roc_lda <- roc(test_set$y, pred_lda)
plot(roc_qda)
plot(roc_lda, add=TRUE, col=2)
```

We can also compare to kin

```{r}
fit <- knn3(y~., data = train_set, k=5)
pred_knn_5 <- predict(fit, newdata = test_set)[,2]
roc_knn_5 <- roc(test_set$y, pred_knn_5)
plot(roc_qda)
plot(roc_knn_5, add=TRUE, col=3)

fit <- knn3(y~., data = train_set, k=51)
pred_knn_51 <- predict(fit, newdata = test_set)[,2]
roc_knn_51 <- roc(test_set$y, pred_knn_51)
plot(roc_qda)
plot(roc_knn_51, add=TRUE, col=4)
```

## Three classes

```{r, echo=FALSE, cache=TRUE}
##Get the truth
dat <- digits %>% filter(label%in%c(1,2,7))
dat <- mutate(dat, label =  as.character(label)) 
row_column <- expand.grid(row=1:28, col=1:28)
ind1 <- which(row_column$col <= 14 & row_column$row <=14)
ind2 <- which(row_column$col > 14 & row_column$row > 14)
ind <- c(ind1,ind2)
X <- as.matrix(dat[,-1])
X <- X>200
X1 <- rowSums(X[,ind1])/rowSums(X)
X2 <- rowSums(X[,ind2])/rowSums(X)
dat <- mutate(dat, X_1 = X1, X_2 = X2)

y <- as.factor(dat$label)
x <- cbind(X1, X2)
library(caret)
fit <- knn3(x, y, 51)
GS <- 150
X1s <- seq(min(X1),max(X1),len=GS)
X2s <- seq(min(X2),max(X2),len=GS)
true_f <- expand.grid(X_1=X1s, X_2=X2s)
yhat <- predict(fit, newdata = true_f, type="prob")
f1 <- loess(yhat[,1]~X_1*X_2, data=true_f,degree=1, span=1/5)$fitted
f2 <- loess(yhat[,2]~X_1*X_2, data=true_f,degree=1, span=1/5)$fitted
f7 <- loess(yhat[,3]~X_1*X_2, data=true_f,degree=1, span=1/5)$fitted
true_f <- true_f %>% mutate(f1=pmin(f1,1), f2=f2, f7=f7) 
rm(dat27,X,X1,X2,fit,GS,X1s,X2s,yhat,f1, f2, f7)

p1 <- true_f %>%
  ggplot(aes(X_1, X_2, fill=f1))  +
  geom_raster()  + #guides(fill=FALSE) +  
  stat_contour(aes(x=X_1,y=X_2,z=f1), 
               data=true_f, breaks=c(0.5),color="black",lwd=1.5) + 
  guides(fill=FALSE)
p2 <- true_f %>%
  ggplot(aes(X_1, X_2, fill=f2))  +
  geom_raster()  + #guides(fill=FALSE) +  
  stat_contour(aes(x=X_1,y=X_2,z=f2), 
               data=true_f, breaks=c(0.5),color="black",lwd=1.5) +
  guides(fill=FALSE)
p3 <- true_f %>%
  ggplot(aes(X_1, X_2, fill=f7))  +
   geom_raster()  + #guides(fill=FALSE) +  
  stat_contour(aes(x=X_1,y=X_2,z=f7), 
               data=true_f, breaks=c(0.5),color="black",lwd=1.5) + 
  guides(fill=FALSE)
library(gridExtra)
grid.arrange(p1,p2,p3, nrow=1)
```

```{r}
##create the training set
set.seed(1)
dat <- sample_n(dat, 3000) %>% select(label, X_1, X_2) %>% mutate(label=as.factor(label))
library(caret)
inTrain <- createDataPartition(y = dat$label, p=0.5)
train_set <- slice(dat, inTrain$Resample1)
test_set <- slice(dat, -inTrain$Resample1)
```


```{r}
train_set %>% ggplot(aes(X_1,X_2,fill=label)) + geom_point(cex=5, pch=21)
```


#### LDA

```{r}
params <- train_set %>% group_by(label) %>% 
  summarize(avg_1 = mean(X_1), avg_2 = mean(X_2), sd_1= sd(X_1), sd_2 = sd(X_2), r = cor(X_1,X_2))
params <-params %>% mutate(sd_1 = mean(sd_1), sd_2=mean(sd_1), r=mean(r))
params 
```

This defines the following estimate of $f(x_1, x_2)$  and the boundary becomes linear:

```{r}
library(mvtnorm)
p0 <- get_p(params[1,], true_f)  
p1 <- get_p(params[2,], true_f)  
p2 <- get_p(params[3,], true_f)  

pred <- apply(cbind(p0, p1, p2),1,which.max)
tmp <- true_f %>% mutate(pred=pred)
tmp %>% ggplot() +
  stat_contour(aes(x=X_1,y=X_2,z=pred),
               breaks=c(1,2,3),color="black",lwd=1.5) +
  geom_point(aes(X_1,X_2,fill=label), dat=test_set,pch=21)
```

#### QDA

```{r}
params <- train_set %>% group_by(label) %>% 
  summarize(avg_1 = mean(X_1), avg_2 = mean(X_2), sd_1= sd(X_1), sd_2 = sd(X_2), r = cor(X_1,X_2))

```

This defines the following estimate of $f(x_1, x_2)$  and the boundary becomes linear:

```{r}
library(mvtnorm)
p0 <- get_p(params[1,], true_f)  
p1 <- get_p(params[2,], true_f)  
p2 <- get_p(params[3,], true_f)  

pred <- apply(cbind(p0, p1, p2),1,which.max)
tmp <- true_f %>% mutate(pred=pred)
tmp %>% ggplot() +
  stat_contour(aes(x=X_1,y=X_2,z=pred),
               breaks=c(1,2,3),color="black",lwd=1.5) +
  geom_point(aes(X_1,X_2,fill=label), dat=test_set,pch=21)
```


#### GLM
```{r}
library(caret)

fit1 <- glm(y~X_1+X_2, data=mutate(train_set,
                                   y=label=="1"),family="binomial")
fit2 <- glm(y~X_1+X_2, data=mutate(train_set,
                                   y=label=="2"),family="binomial")
fit7 <- glm(y~X_1+X_2, data=mutate(train_set,
                                   y=label=="7"),family="binomial")

f_hat1 <- predict(fit1, newdata = true_f, type = "response")
f_hat2 <- predict(fit2, newdata = true_f, type ="response")
f_hat7 <- predict(fit7, newdata = true_f, type = "response")

pred <- apply(cbind(f_hat1, f_hat2, f_hat7),1,which.max)
tmp <- true_f %>% mutate(pred=pred)
tmp %>% ggplot() +
  stat_contour(aes(x=X_1,y=X_2,z=pred),
               breaks=c(1,2,3),color="black",lwd=1.5) +
  geom_point(aes(X_1,X_2,fill=label), dat=test_set,pch=21)
```


#### kNN
```{r}
library(caret)

fit <- knn3(label~., data=train_set, k=51)
f_hat <- predict(fit, newdata = true_f)
f_hat_max <- apply(f_hat,1,max)
pred <- apply(f_hat,1,which.max)
tmp <- true_f %>% mutate(pred=pred)
tmp %>% ggplot() +
  stat_contour(aes(x=X_1,y=X_2,z=pred),
               breaks=c(1,2,3),color="black",lwd=1.5) +
  geom_point(aes(X_1,X_2,fill=label), dat=test_set,pch=21)
```



```{r}
library(caret)

fit <- knn3(label~., data=train_set, k=101)
f_hat <- predict(fit, newdata = true_f)
f_hat_max <- apply(f_hat,1,max)
pred <- apply(f_hat,1,which.max)
tmp <- true_f %>% mutate(pred=pred)
tmp %>% ggplot() +
  stat_contour(aes(x=X_1,y=X_2,z=pred),
               breaks=c(1,2,3),color="black",lwd=1.5) +
  geom_point(aes(X_1,X_2,fill=label), dat=test_set,pch=21)
```






#### Comparison

```{r}
##QDA
params <- train_set %>% group_by(label) %>% 
  summarize(avg_1 = mean(X_1), avg_2 = mean(X_2), sd_1= sd(X_1), sd_2 = sd(X_2), r = cor(X_1,X_2))
p0 <- get_p(params[1,], test_set)  
p1 <- get_p(params[2,], test_set)  
p2 <- get_p(params[3,], test_set) 

pred_qda <- apply(cbind(p0, p1, p2),1,which.max)

##LDA
params <-params %>% mutate(sd_1 = mean(sd_1), sd_2=mean(sd_1), r=mean(r))
p0 <- get_p(params[1,], test_set)  
p1 <- get_p(params[2,], test_set)  
p2 <- get_p(params[3,], test_set)  
pred_lda <- apply(cbind(p0, p1, p2),1,which.max)

##GLM
fit1 <- glm(y~X_1+X_2, data=mutate(train_set,
                                   y=label=="1"),family="binomial")
fit2 <- glm(y~X_1+X_2, data=mutate(train_set,
                                   y=label=="2"),family="binomial")
fit7 <- glm(y~X_1+X_2, data=mutate(train_set,
                                   y=label=="7"),family="binomial")

f_hat1 <- predict(fit1, newdata = test_set, type = "response")
f_hat2 <- predict(fit2, newdata = test_set, type ="response")
f_hat7 <- predict(fit7, newdata = test_set, type = "response")

pred_glm <- apply(cbind(f_hat1, f_hat2, f_hat7),1,which.max)

library(caret)

##KNN 51
fit <- knn3(label~., data=train_set, k=51)
f_hat <- predict(fit, newdata = test_set)
pred_knn_51 <- apply(f_hat,1,which.max)

##KNN 101
fit <- knn3(label~., data=train_set, k=101)
f_hat <- predict(fit, newdata = test_set)
pred_knn_101 <- apply(f_hat,1,which.max)
```

Let's compare:
```{r}
tab <- table(factor(pred_lda, labels=c("1","2","7")), test_set$label)
confusionMatrix(tab)
confusionMatrix(tab)$overall[1]

tab <- table(factor(pred_qda, labels=c("1","2","7")), test_set$label)
confusionMatrix(tab)$overall[1]

tab <- table(factor(pred_glm, labels=c("1","2","7")), test_set$label)
confusionMatrix(tab)$overall[1]


tab <- table(factor(pred_knn_51, labels=c("1","2","7")), test_set$label)
confusionMatrix(tab)$overall[1]


tab <- table(factor(pred_knn_101, labels=c("1","2","7")), test_set$label)
confusionMatrix(tab)$overall[1]
```







