---
layout: page
title: Matrices in R
---

```{r}
x_1 <- c(1,0,0,1)
x_2 <- c(2,4,6,3)
x_3 <- c(6,4,2,4)
X <- cbind(x_1, x_2, x_3)
X
```

Alternatively

```{r}
X <- matrix( c(x_1, x_2, x_3), nrow=4, ncol=3)
X
```

We can extract specific entries using the `[` character:

```{r}
X[2,3]
```

We can extract columns like this:

```{r}
X[,c(1,2)]
```

and rows like this:

```{r}
X[c(2,3),]
```

Be aware that if you select one column or one row, the result is no longer a matrix but a numeric.

```{r}
X[,1]
class(X[,2:3])
class(X[,1])
```
However, we can preserve the matrix class by using the argument `drop`

```{r}
X[,1,drop=FALSE]
```


Useful function:

```{r}
rowMeans(X)
```

```{r}
colMeans(X)
```

Also in the matrixStats package:

```{r}
library(matrixStats)
colSds(X)
colRanks(X)
```

and many other useful functions.

We can apply any function to the rows using apply

```{r}
apply(X,1,mean) ##same as rowMeans
```

or the columns


```{r}
apply(X,2,mean) ##same as rowMeans
```

We can also define our own

```{r}
apply(X,2,function(x){
  c(sum(x[1:2]), sum(x[3:4]))
})
```

One very powerful approach  in it's simplicty and speed is to filter columns based on statistics of those columns:

```{r}
X[ , colMeans(X)>3]
```

We can also do thigs like this:

```{r}
X[X<1] <- 0.5
X
```

or like this

```{r}
X[2,3]<-NA
X[is.na(X)] <- 0
X
```

We can also use this nice trick:

```{r}
X - rowMeans(X)
```

we can scale each row like this:

```{r}
(X - rowMeans(X)) /sd(X)
```

but be careful, it does not work for columns. for columns do this

```{r}
sweep(X, 2, colMeans(X))
```
or
```{r}
t( t(X) - colMeans(X) )
```


Finally, if you know linear algebra, you the transpose:
```{r}
t(X)
```

matrix multiplication:
```{r}
t(X) %*% X
```

crossproduct

```{r}
crossprod(X)
```

inverse
```{r}
solve( crossprod(X))
```

QR decomposition

```{r}
qr(X)
```



