---
title: "Poll Aggregators"
output: html_document
---

## In-class questions
We're going to try something new today. If you have small technical questions during class, go to this [link](https://docs.google.com/document/d/1rClyG2ZMoTM9tER9zjXCBT-26lrpuIH_oSkCLCI6pII/edit?usp=sharing) and ask away.


## Averaging polls


```{r, echo=FALSE, cache=TRUE, warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(pollstR)

theme_set(theme_bw())

race2012 <- pollstr_polls(topic = '2012-president', after= as.Date("2012-11-3"), max_pages = Inf)

polls <- race2012$questions %>% 
  filter(topic=="2012-president" & state=="US") %>% 
  select(choice, value, margin_of_error, observations, id) %>% 
  filter(choice %in% c("Obama","Romney") & !is.na(margin_of_error)) %>% 
  spread(choice, value) %>% 
  left_join(select(race2012$polls, id, pollster, method), by="id") %>%
  filter(method!="Internet") %>%
  mutate(diff= Obama - Romney) %>%
  select(pollster, diff, Obama, Romney, margin_of_error, observations)

arrange(polls,diff) %>% rename( n=observations) %>% 
  mutate(pollster=ifelse(pollster=="PPP (D-Americans United for Change)","PPP",pollster))
```


Given that so many polls pointed to a toss up, why was Nate Silver so confident?


## Competition Data

Let's start by looking at our competition data.

```{r}
library(readr)
filename <- "https://raw.githubusercontent.com/datasciencelabs/data/master/blue-bead-comp-results.csv"
tab <- read_csv(filename)
names(tab)<-c("timestamp", "name", "estimate","N","ci")
```

Looking at the data, are there more blue or red balls?

BTW, the winners are:

```{r}
date <- sapply(strsplit(tab$timestamp," "), function(x)x[1])
date <- as.Date(date,"%m/%d/%Y")
tab %>% mutate(date = date) %>% filter(date<"2016-02-23") %>%
  mutate(diff = abs(estimate - 53.4)) %>% arrange(diff) 
```

Now back to aggregating:

```{r}
tab <- mutate(tab,estimate=ifelse(estimate<1, estimate*100, estimate))
boxplot(tab$estimate)

tab = filter(tab, estimate>20)
```


So now it's my turn to enter a competing entry. 

We can assume that all these entries were based on independent data. So this data actually shows data for

```{r}
tab %>% summarise(sum(N))
```

We can deduce how many blue and red beads the first person saw:

```{r}
slice(tab,1)
```

In general we know each person saw

```{r}
tab %>% mutate(blue = estimate/100 * N, red = (1-estimate/100) * N) %>% select(blue, red)
```

So we can add these all up and create an estimate on thousands of draws:

We have $k=1,\dots,K$ polls each with $N_k$ observations and en estimate $\hat{p}_k$. We can deduce then that for poll $k$, $N_k \hat{p}_k$ are blue. So the aggregate $\hat{p}$ is

$$
\frac{\sum_{k=1}^K \hat{p}_k N_k}{\sum_{k=1}^K N_k}
$$
which can be viewed as a weighted average:

$$
\sum_{k=1}^P w_k \hat{p}_k \mbox{ with } w_k = \frac{N_k}{\sum_{k=1}^K N_k}
$$

So our new estimate is

```{r}
N <- sum(tab$N)
p_hat <- sum(tab$estimate*tab$N)/N
```

And our confidence interval is pretty tight:
```{r}
c(-1,1)*qnorm(0.975)*sqrt(p_hat/100*(1-p_hat/100))/sqrt(N)*100
```

## Back to the 2012 Election

Let's compute this for Obama and Romney (we assume the undecided at random)

```{r}
O <- sum(polls$Obama*polls$observations)/sum(polls$observations)/100
R <- sum(polls$Romney*polls$observations)/sum(polls$observations)/100
```

and the difference is 
```{r}
round(O-R,2)
```

The margin of error for each one is:

```{r}
N <- sum(polls$observations)
round( qnorm(.975)*sqrt(O*(1-O))/sqrt(N), 3)
round( qnorm(.975)*sqrt(R*(1-R))/sqrt(N), 3)
```

Now we have two confidence intervals?  We can form a confidence interval for the difference. The difference is approximately $\hat{p} - (1-\hat{p})$ or $1-2\hat{p}$. This implies that the standard error for the difference is twice as large than for $\hat{p}$. So our confidence interval is:

```{r}
O-R + c(-2,2)*qnorm(.975)*sqrt(O*(1-O))/sqrt(N)
```

So a 95% confidence interval still didn't quite call the election for Obama (but it was close). However, as we will learn there was much more involved in the "90% chance of winning"" prediction which we will learn next.

## More Polls
 
Let's get polls for two months before. 

####Assessment: 
Adapt the code above to make a plot of difference between Obama and Romney against time.



# Questions?
If you have any questions, please fill this out: http://goo.gl/forms/BWzC2kVNZC
