---
title: "Predicting Elections"
output: html_document
---

## In-class questions: 3/7/2016

If you have small technical questions during class, go to this 
[link](https://docs.google.com/document/d/1Lz0NsFsgE9ZgKnIQrJp-kt2d3rx0xWRZB23-pdZpsfY/edit?usp=sharing) and ask away.


## 2008 Presidential Election (Obama vs McCain)

We are going to put together the concepts we have
learned related to modeling polls. For this, we will
continue to use the 2008 presidential election 
as an example. We have previously shown how to scrape
the data from 
[this](http://www.pollster.com/08USPresGEMvO-2.html) web site.

```{r}
library(XML)
theurl <- paste0("http://www.pollster.com/08USPresGEMvO-2.html")
html_tables <- readHTMLTable(theurl,stringsAsFactors=FALSE)
tab <- html_tables[[1]]
```


## Wrangling the data

We have already discussed in detail how to scrape and wrangle
this particular data set in the  
[models.Rmd](https://github.com/datasciencelabs/2016/blob/master/lectures/models/models.Rmd) 
in the 2016 GitHub repository. We used the `tidyr`, `stringr` and 
`lubridate` packages. Here, I have combined all the steps from the 
[models.Rmd](https://github.com/datasciencelabs/2016/blob/master/lectures/models/models.Rmd) 
into one step. 

```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(stringr)
library(lubridate)

tab <- tbl_df(tab)
tab <- tab %>% 
            separate(col=Dates, into=c("start_date","end_date"), 
                     sep="-", fill="right") %>% 
            mutate(end_date = ifelse(is.na(end_date), start_date, end_date)) %>% 
            separate(start_date, c("smonth", "sday", "syear"), sep = "/", 
                    convert = TRUE, fill = "right") %>% 
            mutate(end_date = ifelse(str_count(end_date, "/") == 1, 
                                     paste(smonth, end_date, sep = "/"), end_date)) %>%
            mutate(end_date = mdy(end_date)) %>% 
            mutate(syear = ifelse(is.na(syear), year(end_date), syear + 2000)) %>% 
            unite(start_date, smonth, sday, syear) %>% 
            mutate(start_date = mdy(start_date)) %>% 
            select(-Barr, -Nader, -Other, -Undecided, -Margin) %>% 
            separate(`N/Pop`, into=c("N","population_type"), sep=" ", 
                convert=TRUE, fill="left") %>% 
            mutate(Obama = as.numeric(Obama)/100, 
              McCain=as.numeric(McCain)/100,
              diff = Obama - McCain,
              day=as.numeric(start_date - mdy("11/04/2008")),
              week = floor(day/7))
```


## Using Hierarchical Model to Predict Polls

Last week, we introduced hierarchical models as a tool to estimate 
the posterior distribution of the [batting average of Jose Iglesias](https://github.com/datasciencelabs/2016/blob/master/lectures/models/bayes.Rpres). 

As a brief summary of using hierarchical models, 
you build a model before seeing polls (the prior distribution):  
$$
\theta \sim N(\mu, \tau^2) 
$$

The you start seeing poll data (the sampling distribution):

$$
Y | \theta \sim N(\theta, \sigma^2) 
$$

We update our prediction (the posterior distribution):

$$
\begin{eqnarray*}
\mbox{E}(\theta|y) &=&  \mu + (1-B)(Y-\mu)
\mbox{ with }B &=& \frac{\sigma^2}{\sigma^2+\tau^2}\\
\\
\mbox{var}(\theta\mid y) &=& \frac{1}{1/\sigma^2+1/\tau^2}
\end{eqnarray*}
$$

That is a lot information stuffed into three steps. 
Let's break the steps down individually and apply the theory to 
the 2008 Presidential Election poll data to see how it works.


### Prior belief

The first step is to define our *prior belief* that Obama will beat
McCain. We define a model (*a prior*) before seeing polls for the 
difference between `Obama` and `McCain` (which we define 
as $\theta$):

$$
\theta \sim N(\mu, \tau^2) 
$$

To quantify this belief, we must define values for $\mu$
and $\tau$. 

#### Assessment 
Say before seeing any poll data, you have no idea who is going to win.  
What should $\mu$ be? What about $\tau$? 

**Your answer here**:


Now, let's assume we are part of the republican party and we 
passionately believe that McCain will beat Obama. To quantify 
this belief, we can say the distribution of $\theta$
is normal with mean $\mu = -0.15$ and standard deviation
$\tau = 5$.

#### Assessment 
Under your prior belief that McCain will beat Obama, 
what is the probability that Obama will beat 
McCain in the 2008 presidential election? 

```{r}
##Your code here

```


### Observed data (sampling distribution)

[Last week, we learned about how we have been implicitly using linear models in this course](https://github.com/datasciencelabs/2016/blob/master/lectures/inference/models.pdf). 
We have been doing this by trying to estimate the probability of tossing 
a coin and observing a head, the proportion of blue beads in a jar, 
the average height of a population, and the difference in the 
proportion of votes two candidates, for example Obama and McCain, 
will receive. The goal of using linear models is to try and estimate 
some parameter $\theta$ by observing draws from a sampling model.  
The first linear model we considered was: 

$$ Y_i = \theta + \varepsilon_i, i = 1, ..., N$$

where $Y_1, Y_2, \ldots, Y_N$ are the observed values with $N$ 
total observations. Each observation has an *error* term $\varepsilon_i$
that defines the difference between the true parameter value ($\theta$)
and the observed value ($Y_i$). We assume that the expected value 
of the error term is 0 and very often assume the 
standard deviation is a contant $\sigma$. 

Up until now, we have not used the poll data. We have 
only made statements about our prior beliefs. 
Now assume you start seeing the poll data. Let $Y$ be the 
observed difference in the polls between Obama and McCain. 
We define the sampling distribution as

$$
Y | \theta \sim N(\theta, \sigma^2/N) 
$$

where 

* $Y$ is the observed difference in each poll between Obama and McCain
* $\theta$ is the true difference in the polls on election night
* $\sigma^2$ is the observed sample variance
* $N$ is the number of polls we observe (i.e. each poll is an observation here)

#### Estimate $\theta$

Using our linear model defined above, we have seen how to use the 
method called 
[*least squares*](https://en.wikipedia.org/wiki/Least_squares), 
which minimizes the least squares equation: 

$$ \sum_{i=1}^N (Y_i - \theta)^2 $$

to obtain the *least squares estimate* (LSE), which is basically a
sample average of the individual observed polls: 

$$ \hat{\theta} = \frac{1}{N}\sum_{i=1}^N Y_i = \bar{Y} $$

Because this is a sample average from a sampling model, 
we know its expectation is $\theta$ and it’s standard error is 
$\sigma / N$ where $N$ is the number of observations 
and $\sigma$ the standard deviation of the distribution of $\varepsilon$. 

Now, what is $\sigma$? 

#### Estimate $\sigma$ using Central Limit Theorem (theoretical) versus the sample standard deviation (empirical)

We have learned that in the context of estimating the difference 
in the polls between two presidentical candidates, we can use the 
Central Limit Theorem (CLT) to model $\varepsilon$. According to the CLT,

$$ \varepsilon \sim \mbox{Normal}\left(0, \frac{2\sqrt{p (1-p)}}{\sqrt{n}}\right) $$

where $p$ is the prorportion for Obama, $n$ is the poll size from a single 
poll. This means statistical
theory tells us the expected value of $\varepsilon$ is 0 and the 
standard deviation $\sigma$ should be given by
$\frac{2\sqrt{p (1-p)}}{\sqrt{n}}$. 

Remember, if $p$ represented the percentage for just one 
candidate (e.g. Obama), the statitical theory tells us that the 
standard error for the estimate of the percentage for Obama is 

$$ \frac{\sqrt{p (1-p)}}{\sqrt{n}} $$

But if we are estimating the difference in percentages between two 
candidates, then the estimate is approximated by 
$\hat{\theta} = \hat{p} - (1-\hat{p}) = 2\hat{p} - 1$. 
This implies that the standard error is $2 \sqrt{ p (1 - p)} / \sqrt{n}$.

Because each $Y_i$ a separate poll then the standard deviation of 
$\varepsilon$ should be about $2 \sqrt{ p (1 - p) }/ \sqrt{n}$:

**Note, this theory is extremely useful when we just have one poll.**

**However**, if we are fortunate enough to be able to observe mutliple 
polls results (e.g. polls from different pollsters or polls from 
different weeks over time), we can use the poll data directly 
to estimate $\sigma$ (instead of using the statistical theory above). 
The typical strategy is to use the sample standard deviation formula: 

$$ s = \sqrt{ \frac{1}{N-1}\sum_{i=1}^N (Y_i - \bar{Y})^2} $$

where $N$ is the total number of polls we observe. 

**IMPORTANT**: Do not confuse the poll sample size $n$ with the 
number of observations in our model above $N$. In the model above we
treat each poll as a single observation and we observe $N$ polls. But
it is important to keep in mind that each poll we observe had 
a poll size associated with it (i.e. $n$). 




### Explore the data to estimate $\theta$ and $\sigma$

Next we are going to explore and use the 2008 presidential election 
poll data to try and estimate $\theta$ and $\sigma$. 

#### Assessment 
Assume today is Jan 1, 2008. Use all the poll data 
up until today to provide an estimate for $\theta$ 
and an estimate of the standard deviation $\sigma$
(empirically or from the data and not using 
the statistical theory from the CLT).

```{r}
##Your code here.

```

Last week, we also a scatter plot showing the number of polls 
(y-axis) for each week (x-axis). 

```{r}
library(ggplot2)
tab %>% 
    group_by(week) %>% 
    summarize(num_polls=n()) %>%
    ggplot(aes(week, num_polls)) + geom_point()
```

Many of the polls 50 weeks prior to the election only had a handful of polls. 

#### Assessment 
In which weeks before the election were there 
at least five or more polls in a given week. 

```{r}
##Your code here.

```

#### Assessment 
Assume today is 30 weeks away from the 2008 election. Extract 
the polls that occurred between Jan 1, 2008 and 30 weeks prior
to the election (i.e. Assume we do not trust any polls that 
occurred too far out). Remove polls from weeks that do
not have at least 5 polls. Provide an estimate for $\theta$
called `theta_hat` and an estimate of the standard 
deviation $\sigma$ called `sigma_hat`. In addition
record the number of total number of polls and call 
it `tot_poll`. We will use these estimates for our sampling
distribution. Save your estimates into a table called `tabEst`.

**Hint**: You will need to group the polls by week using 
`group_by()`.  After you have filtered for the poll 
criteria, use `ungroup()` before estimating the 
mean and standard deviation. 


```{r}
##Your code here.

```


#### Assessment 
Use the Central Limit Theorem to construct 
a confidence interval for $\hat{\theta}$ using the empirical
approach to estimate the standard deviation $\sigma$. 

```{r}
##Your code here

```

#### Assessment

Based on our sampling distribution, 
what is the probability our interval includes the 
true difference between Obama and McCain in 
the 2008 presidential election? 

```{r}
##Your code here

```


#### Compare the empirical estimate $\sigma$ to the theoretical estimate of $\sigma$ using Central Limit Theorem

Until now, we have been using the sample standard deviation to estimate 
$\sigma$ (i.e. empirically).  Now let's compare how the empirical estimate
of $\sigma$ compares to the theoretical estimate of $\sigma$ using 
the statistical theory that we've learned. 

Let's filter the polls for only polls within 5 weeks of the 
election and for polls that had a sample size of 1000.  We can use 
boxplots to plot the distribution of the difference between 
Obama and McCain stratified by pollster. 
```{r}
tab %>% 
    filter(week>-5, N == 1000) %>%
    ggplot(aes(Pollster, diff, fill=Pollster)) + geom_boxplot()
```

If you recall above, we discussed how use the statistical 
theory from the Central Limit Theorem (CLT) to estimate 
$\sigma$. According to the CLT, 

$$ \varepsilon \sim \mbox{Normal}\left(0, \frac{2\sqrt{p (1-p)}}{\sqrt{n}}\right) $$

where $n$ is the poll size. This means statistical
theory tells us the expected value of $\varepsilon$ is 0 and the 
standard deviation $\sigma$ should be given by
$\frac{2\sqrt{p (1-p)}}{\sqrt{n}}$.

#### Assessment 
Use the poll data above (i.e. polls within 5 weeks of the 
election and for polls that had a sample size of 1000), 
estimate $\theta$ and $\sigma$ using the statistical 
theory from the CLT. 

```{r}
##Your code here.

```


#### Assessment 
In contrast, use the poll data above (polls within 5 weeks of the 
election and for polls that had a sample size of 1000) to 
estimate $\sigma$ using the sample standard deviation. 

$$ s = \sqrt{ \frac{1}{N-1}\sum_{i=1}^N (Y_i - \bar{Y})^2} $$

How does it compare to the estimate of $\sigma$ above using the CLT? 
```{r}
##Your code here.

```

### Why does the theory not match what we see in the data? 

In general, if the model is a good fit to the data, the 
empirical estimate of $\sigma$ should match what the 
the statistical theory says it should be. Recall this is our model: 

$$ Y_i = \theta + \varepsilon_i, i = 1, ..., N$$

Instead, we see the empirical estimate of $\sigma$ 
is larger than what the theory says it should be.
Why is this?? 

Well, there are many possible reasons for this. 
Let's consider three reasons: 

#### 1. Time effect 
One possible reason is that if there is a strong 
time (or "week") effect, then there is extra variability in the 
data that is not being accounted for in the model. 

```{r}
tab %>%
    ggplot(aes(end_date, diff)) + geom_point() + 
        geom_smooth(span=0.5) + geom_hline(aes(yintercept=0)) + 
        geom_hline(aes(yintercept=0.072))
```

We could modify the model to include another term $w_t$ that 
represents the time (or "week") effect. 

$$ Y_{t,i} = \theta + w_t + \varepsilon_{t,i} $$

**Note**: We now have two indexes $t$ denoting week and
$i$ an index for the $i$-th poll during week $t$.

By including this extra term in our model, we are 
saying that the variability observed in the polls 
does not just come from $\varepsilon$. We are 
saying there is extra or additional variabilty from
time that we are ignoring. 

We can model the time effects $w_t$ by making 
assumptions about the expected value 
and variance of of $w_t$. For example, we could 
model this as a fixed effect and estimate it, for 
example, [loess](https://en.wikipedia.org/wiki/Local_regression). 
This is not that useful for forecasting as we don't 
know if the trend will continue. More useful is 
to model it as a random effect with its variance 
depending on days left to the election (e.g. a model 
with a decreasing variance as the weeks get closer 
to the election.

To incorporate information about the time effect, 
we an use the fact that we see the variance 
in the difference between Obama and McCain decreases 
as we get closer to election night.  

If we are 10 weeks away from election night, you can get an
idea of the standard deviation of $w_t$ by looking at 10 weeks prior on
previous elections. In previous elections we observe $\theta$ so we
can actually observe $w_t$ + $\varepsilon$. We can 
observe the this for several weeks and
notice that it gets smaller and smaller as the election gets closer.


#### 2. House effect 
Another possible reason is from what's called 
a *house effect* or pollster effect. If there was no pollster
effect, we would expect the distributions (or boxplots)
of the difference between Obama and McCain to be similar.  

```{r}
tab %>% 
    filter(week > -4) %>%
    group_by(Pollster) %>% 
    filter(n()>4) %>% 
    ggplot(aes(Pollster, diff , fill=Pollster)) + geom_boxplot()
```

We see that each pollster has a different distribution in the
difference between Obama and McCain. We could modify the model 
to include another term $p_j$ that represents the 
pollster effect. 

$$ Y_{t,i,j} = \theta + w_t + p_{j} + \varepsilon_{t,i,j} $$

where $p_{j}$ a pollster effect for the $j^{th}$ pollster. 
By including this extra term in our model, we are 
saying that the variability observed in the polls 
does not just come from $\varepsilon$. We are 
saying there is extra or additional variabilty from
the pollsters.  

We can model the pollster specific effects $p_j$ by making 
an assumption about the expected value of $p_j$ 
(e.g. $\mbox{E}(p_j)=0$) and we can model the 
pollster specific effects to have different variances. 
To estimate these we can use previous election data. 
With these in place, we can construct weighted estimates
for $\theta$ that better estimate the variablity. 


#### 3. General biases

The last possible reason we'll consider here is from 
what are called *general biases* that have not been 
accounted for. Specifically, our assumption that 
$\mbox{E}(p_j)=0$ is incorrect. This assumption says that,
on average, pollsters are not biased, but this is not the
case. Instead we need to add a general bias to the model

$$Y_{t,i,j} = \theta + w_t + p_j + b + \varepsilon_{t,i,j}.$$

But note we cannot estimate $b$ from the data: this 
model is not identifiable. However, we can model $b$ 
as a random effect with and estimate its variance from 
past elections where we know $\theta$. 


### Posterior distribution

The last step in the hierarchical model is to obtain a 
posterior distribution. This means we will combine 
our prior belief with our sampling distribution to 
update our prediction of Obama beating
McCain. To do this, we will use the formulas we 
learned last week: 

$$
\begin{eqnarray*}
\mbox{E}(\theta|y) &=&  B\mu+ (1-B)Y \\
& & \mbox{ where } B = \frac{1/\tau^2}{N/\sigma^2+1/\tau^2}\\
\\
\mbox{var}(\theta\mid y) &=& (1-B)\sigma^2/N
\end{eqnarray*}
$$

Note that there are lots of equivalent ways to write $B$, 
but this one shows that it is a precision weight, 
based on the ratio of the prior precision to the 
total precision of the prior and the data. 
Precision here is the inverse of the variance. 

Equivalently,
$$
\begin{eqnarray*}
\mbox{E}(\theta|y) &=&  B\mu+ (1-B)Y \\
& & \mbox{ where } B = \frac{1/\tau^2}{N/\sigma^2+1/\tau^2}\\
\\
\mbox{var}(\theta\mid y) &=& \frac{1}{N/\sigma^2+1/\tau^2}
\end{eqnarray*}
$$

#### Assessment

Combine the prior distribution with the sampling distribution to 
estimate the mean and standard deviation of the 
posterior distribution for $\theta$. 

```{r}
##Your code here. 

```

Use the posterior distribution, estimate 
the probability of Obama beating McCain in the 
2008 Presidential election. 

```{r}
##Your code here

```



### Monte Carlo simulation to generate election results

Now we will use the posterior distribution in a Monte Carlo 
simulation to generate election results. In each simulation, 
we will simulate data from the posterior distribution and 
compute the probability of Obama beating McCain. 
This a histogram of these results.

First, let's simulate data from the posterior distribution for one observation. 
Keep in mind, we are simulating the difference between Obama and McCain.
```{r}
## Will discuss in class. 

```

Next, let's simulate data from the posterior distribution for $N$ observations. 
```{r}
## Will discuss in class. 

```

We see that in some cases the difference is below 0, but the distribution
is centered at `mu_post`. We can count the number of times Obama beat
McCain from our observations generated from the posterior distribution. 

```{r}
## Will discuss in class. 

```

Now let's put it all together in a Monte Carlo simuation to generate 
election results. 

```{r}
## Will discuss in class. 

```

#### Assessment
Next, assume there is a general poll bias that 
is a random variable. It can go either way. 
How can you add this to your simulation? 
How does it affect the distribution?

```{r}
##Your code here.

```


